Analyzing DECREASE trials for extent of data fabrication

### 2017-01-26 submission/manuscript.Rmd
Revise up to step 3 methods

---
title: "Analyzing DECREASE trials for extent of data fabrication"
author: "Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

The effectiveness of beta-blockers in preventing perioperative mortatility in non-cardiac surgery has been subject of discussion due to findings of misconduct. [@commissie2011;@commissie2012;@commissie2013;@bouri2014]. Perioperative mortality is the deathrate of patients during the period of the surgical procedure, typically including admission, anaesthesia, surgery, and recovery.

The trials that were found to be suspect were the DECREASE trials [@commissie2011;@commissie2012;@commissie2013]. Upon being discovered as (potentially) fabricated, the DECREASE studies were excluded from a 2014 meta-analysis [@bouri2014]. The meta-analyses restricted to non-DECREASE studies found that conclusions on the effectiveness of beta-blockers in perioperative mortality were reversed --- beta-blockers seemed to actually *increase* perioperative mortality instead of decreasing it.

The committee reports about the DECREASE studies indicate that data fabrication was likely [@commissie2011;@commissie2012;@commissie2013], but the extent of the data fabrication remained unclear. In this paper, we aim to estimate the extent of data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999].

To investigate the evidence for data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. First, we tried to replicate the findings from the 2014 meta-analysis [@bouri2014]. Second, we evaluated the veracity of the DECREASE studies (i.e., the probability that these study findings occurred assuming no data fabrication). Third, we reversed this assumption and assumed that data fabrication did occur, estimating how many data points would have to be fabricated to reproduce the results of the DECREASE studies.

# Step 1: reproducing Bouri et al. (2014)

@@

To ensure that we were using similar analytic procedures as in the 2014 meta-analysis [@bouri2014], we initially reproduce Bouri et al.'s estimates. This ensured that (1) their results are reproducible and (2) we are using the correct estimates in the following steps. Using figures 2 and 3 from the original paper[@bouri2014], we extracted the raw event data for the 2 (control vs experimental) by 2 (event or no event) design, which we used to recompute the natural logarithm of the risk ratio and its standard error. The extracted event data is available at [osf.io/aykeh](osf.io/aykeh) and our analysis plan was preregistered at [osf.io/vnmzc](osf.io/vnmzc).

```{r, echo = FALSE}
suppressPackageStartupMessages(if(!require(metafor)){install.package('metafor')})
suppressPackageStartupMessages(library(metafor))
suppressPackageStartupMessages(if(!require(latex2exp)){install.package('latex2exp')})
suppressPackageStartupMessages(library(latex2exp))

bouri_data <- read.csv('../data/bouri_data.csv')
# bouri_data <- read.csv('data/bouri_data.csv')

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled these using the `R` package `metafor` [@viechtbauer2010]. We estimated a weighted random-effects model using the restricted maximum-likelihood estimator (i.e., `REML`)[@viechtbauer2005] to estimate the variance of effects. We used the default weighting procedure in the `metafor` package. When there was a zero-count for a cell, .5 was added to each cell of that trial [@agresti2002]. The 2014 meta-analysis [@bouri2014] did not specify the effect variance estimate used. Hence, minor discrepancies between our estimates and the original estimates might be due to differences in the estimation procedure.

@@

First, we analyzed the RRs to verify the estimates in Figure 2 and found that we were able to reproduce the estimates for the different sets of studies. The original Figure 2 [@bouri2014] differentiated between the estimates from the non-DECREASE trials and the DECREASE trials. We confirmed the estimates of the effect size and the variance for both the non-DECREASE- and the DECREASE trials, save for some minor discrepancies due to the estimation method. Table `r tab <- 1; tab` depicts the original and reproduced values for both sets of studies.

*Table `r tab`*. The original- and reproduced meta-analytic results based on the data provided in the 2014 meta-analysis.[@bouri2014]

|              |            | Risk ratio   | $\tau^2$      | Confidence interval |
|--------------|------------|---------------------|-------------------|---|
| Non-DECREASE | Original   | 1.27                   | 0.00                 | [1.01; 1.60] |
|              | Reproduced | `r round(exp(mod_clean$b), 2)`  | `r round(mod_clean$tau2, 2)`                 | `r sprintf('[%s; %s]', round(exp(mod_clean$ci.lb), 2), round(exp(mod_clean$ci.ub), 2))` |
| DECREASE     | Original   | 0.42                   | 0.29                 | [0.15; 1.23] |
|              | Reproduced | `r round(exp(mod_dirty$b), 2)` | `r round(mod_dirty$tau2, 2)` | `r sprintf('[%s; %s]', round(exp(mod_dirty$ci.lb), 2), round(exp(mod_dirty$ci.ub), 2))` |

@@

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to replicate results presented in Figure 4 of the 2014 meta-analysis.[@bouri2014] Surprisingly, our results showed more evidence against equal subgroups than the original meta-analysis [@bouri2014]. Whereas the original test for subgroup differences was $\chi^2(1)=3.91,p=.05$, our analyses indicated that $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$. Additionally, the original analyses showed substantial heterogeneity ($I^2=74.4$\%) where we found an $I^2=`r round(mod_mods$I2, 3)`$\%. Different estimates of the heterogeneity (e.g., `DerSimonian-Laird` instead of `REML`) did not change this difference.

<!--- Emailed bouri about this on 2016 03 16--->

@@

We estimated the probability of obtaining the results in the DECREASE studies based on the estimates from the non-DECREASE studies. To this end, we assumed that the non-DECREASE studies provide a valid representation of the true effect of beta-blockers on perioperative mortality. The estimated probability is also known as the veracity of the data [@peters2015], which indicates the probability of the observed data under a given true effect. We assumed that the non-DECREASE studies estimate the true population distribution of effects, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset.

@@

Based on the estimated mean log RR and credible interval in the non-DECREASE studies, we computed the probability of log RR in the DECREASE trials. The estimates of the non-DECREASE studies were obtained from step 1, which include the estimated log RR (i.e., `r round(mod_clean$b, 2)`), and the 95\% credibility interval as provided by the package `metafor` (i.e., `r sprintf('[%s; %s]', round(mod_clean_pred$cr.lb, 3), round(mod_clean_pred$cr.ub, 3))`. We assumed a normal distribution of population effects with the estimated effect as the mean of the distribution. The 95\% credibility interval denotes the bounds of the normal distribution that covers 95\% of the density, where the standard deviation is calculated as the distance from the mean to either bound, divided by 1.96. This allows for an approximation of the population effect distribution, as depicted in Figure `r fig <- 1; fig`.

@@

# pdf('../figures/fig1', width = 8, height = 6)
# use http://pdf2png.com/ when preparing manuscript
# for high quality plots
par(mar = c(.5,0,0,0))
plot(x = temp,
     dnorm(x = temp,
           mod_clean$b,
           (mod_clean_pred$pred - mod_clean_pred$cr.lb)/1.96),
     type = 'l',
     bty = 'n',
     xaxt = 'n',
     yaxt = 'n',
     xlab = '',
     ylab = '')

@@

# dev.off()
```

*Figure `r fig`*. Density plot of the estimated true effect distribution based on the non-DECREASE studies, with the position of the DECREASE studies highlighted.

@@

Figure `r fig` indicates that the DECREASE trials are highly unlikely under the estimated effect distribution based on the non-DECREASE trials. More specifically, DECREASE-I (or more extreme) has a probability of $`r prob[1]`$ (less than 1 in a quintillion) and DECREASE-IV $`r prob[2]`$ (3 in a billion), which indicates they are unlikely to come from the same population effect distribution. Observing two of such extremely unlikely results in this population effect distribution is highly improbable, $`r prob[1]*prob[2]`$. This would therefore indicate that the DECREASE trials are severely different from the non-DECREASE studies.

We note that the variance in the non-DECREASE studies was estimated at 0, indicating that there would supposedly be a fixed-effect of beta-blockers on perioperative mortality despite differences in type of beta-blockers administered and dosage/duration of the beta-blockers. This homogeneity is not excessive, considering that the *Q*-statistic is sufficiently large to begin with (i.e., `r round(mod_clean$QE, .2)`).[@ioannidis2006] Nonetheless, the original non-DECREASE trials show substantive variation in the treatment, for example the usage of different beta-blockers (i.e., atenolol, propanolol, bisoprolol, metoprolol) in different (standardized) dosages, which have been argued to affect effectiveness of the interventions. As such, the lack of heterogeneity can be genuine and indicative of no difference between these different beta-blockers and their dosage/duration. Nonetheless, it could also be due to uncertainty in the estimate of heterogeneity $\tau^2$ due to the small set of studies included.

Considering that the variance (i.e., heterogeneity) estimate is 0 and uncertain due to the small number of studies included, sensitivity analyses of the results are in place. The probability of observing both DECREASE trials stays approximately below 1 out of 1000 until the variance estimate is .25, and the probability stays approximately below 1 out of 100 until the variance estimate is .43 (see Figure `r fig <- fig + 1; fig`). Further research would be worthwhile to better estimate the variance in the effects of beta-blockers on perioperative mortality.

```{r, echo = FALSE}
tau2 <- seq(0, 2, .01)

@@

par(mar = c(4, 4, 0, 0))
plot(tau2, prob_sens_t, type = 'l', xlab = TeX('$\\tau^2$'),
     ylab = 'p-value', bty = 'n')
points(x = tau2[c(26, 44)], y = prob_sens_t[c(26, 44)])
text(x = tau2[c(26, 44)],
     y = prob_sens_t[c(26, 44)] + .02,
     labels = c(TeX('$\\tau^2=.25$'), TeX('$\\tau^2=.43$')),
     cex = .8)
```

*Figure `r fig`*. Sensitivity analyses for the *p*-value that indicates the probability of observing the results from the DECREASE studies, or more extreme results, based on the estimated true effect (non-DECREASE trials) and the accompanying variance estimate.

@@

We estimated the number of data points that would need to be fabricated to arrive at the estimates from the DECREASE studies, given that the non-DECREASE studies represent the true effect. This assumes, in contrast to Step 2, that the DECREASE studies in fact contain fabrication data. The estimates from Step 3 provide an indication of the extent of the potential data fabrication in the DECREASE studies under this assumption.[@commissie2011;@commissie2012;@commissie2013;@bouri2014]

@@

In order to estimate the number of fabricated data points, we meta-analyzed the log odds of mortality per condition for both the DECREASE- and non-DECREASE studies separately. For each of these, we ran a meta-analysis applying the same methods as in Step 1. These separate analyses were necessary, because Step 1 only provided estimates of the overall risk ratio, but not of the odds conditional on exposure (beta-blocker or control) and study set (DECREASE- versus non-DECREASE).

We applied the inversion method to estimate the number of data points fabricated in the DECREASE-I and DECREASE-IV studies.[@casella2002] The inversion method iteratively simulates data with *X* fabricated data points; *X* varying from 0 through *N*. For each *X*, 10,000 iterations were conducted. Data fabrication occurred using a binomial distribution, with the population effect from the DECREASE studies serving as the probability of an event occurring. The population effect was either drawn from the estimated population distribution, which introduced uncertainty (i.e., uncertain estimate), or was equated to the point estimate (i.e., certain estimate; not preregistered). The remaining genuine data, *N - X*, were simulated in a similar fashion, where we only used the effect distribution from the non-DECREASE studies. Using the (partially) fabricated dataset, we estimated the risk ratio from the data and noted whether the two-sided *p*-value was larger under the non-DECREASE effect distribution (coded 0) or the DECREASE effect distribution.
<!--- Marcel's Likelihood approach should be incorporated still; DEVIATION FROM PREREGISTRATION --->

Based on the results of the inversion method, we obtain a confidence interval of data points fabricated in the DECREASE studies. After running all iterations for one *X*, the proportion of trials that were more likely under the DECREASE population effect than under the non-DECREASE population effect were computed. This resulted in a proportion for each *X* from 0 to *N*. Those proportions closest to .025 and .975 denoted the 95\% confidence interval of data points fabricated.

@@

par(mar = c(2.5, 5, 0.5, 0.5),
    mfrow = c(2, 2))
plot(res_i_uncertain,
     xlab = '',
     ylab = 'Proportion',
     ylim = 0:1,
     type = 'l',
     lty = 1)
lines(res_i_certain, pch = 2, type = 'l', lty = 2)
abline(h = c(.025, .975), lty = 3)

@@

@@

par(mar = c(4, 5, 0.5, 0.5))
plot(res_iv_uncertain,
     xlab = 'Data-points fabricated',
     ylab = 'Proportion',
     ylim = 0:1,
     type = 'l',
     lty = 1)
lines(res_iv_certain, pch = 2, type = 'l',
      lty = 2)
abline(h = c(.025, .975), lty = 3)

plot(x = NA, y = NA,
     xlim = c(0,1), ylim = c(0, length(res_iv_certain)),
     xlab = "Confidence in %", ylab = "Data points fabricated")

lines(x = seq(0, 1, .01), y = res_iv_uncertain_lb)
lines(x = seq(0, 1, .01), y = res_iv_uncertain_ub)
lines(x = seq(0, 1, .01), y = res_iv_certain_lb, lty = 2, pch = 2, type = 'l')
lines(x = seq(0, 1, .01), y = res_iv_certain_ub, lty = 2, pch = 2, type = 'l')
abline(v = .95, lty = 3)
```
*Figure `r fig <- fig + 1; fig`*.

>>>>

---
title: "Analyzing DECREASE trials for extent of data fabrication"
author: "Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    keep_tex: yes
  word_document: default
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

<!-- Introduction
What should be in here?
1. Effectiveness of beta-blockers (decrease probability))
2. Effects of findings of misconduct
3. Why this study. -->
The effect of beta-blockers on perioperative mortality in non-cardiac surgery for patients with coronary artery disease (CAD) has been subject of discussion [@Coleg5210] due to findings of research  misconduct [@commissie2011;@commissie2012;@commissie2013] that were central to claims of their effectiveness [@Devereaux313;@Angeli2010;@bouri2014]. Perioperative mortality is the deathrate of patients during the period of the surgical procedure, which typically includes admission, anaesthesia, surgery, and recovery. This related finding of research misconduct altered meta-analytic results of perioperative mortality in non-cardiac surgery for CAD patients substantively[@bouri2014], where meta-analyses that include these studies conclude that beta-blockers decrease perioperative mortality [@Devereaux313;@Angeli2010;@bouri2014] but a meta-analysis that excludes these studies concludes that beta-blockers increase perioperative mortality[@bouri2014].

The trials deemed to be the result of research misconduct were the Dutch DECREASE trials[@commissie2011;@commissie2012;@commissie2013]. The committees that investigated the integrity of the DECREASE studies reported that data fabrication was likely [@commissie2011;@commissie2012;@commissie2013], but that the extent of the data fabrication remained unclear. Given that the inclusion or exclusion of these trials in meta-analyses is crucial in determining whether beta-blockers are indeed effective during the perioperative period and the latest guidelines still recommend the usage of beta-blockers [@esc2014], we aim to estimate the extent of data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999] in order to further this debate.

The reports on the DECREASE trials primarily focused on the provenance of the raw data and patient files (e.g., informed consent, whether data corresponded to patient files), but neglected the extent to which the DECREASE studies deviated from similar trials that tested the effectiveness of beta-blockers on perioperative mortality. Comparing how (dis)similar trials are from comparable studies is not a new idea [@Buyse1999-jq] and has proven effective in detecting data fabrication [@Knepper2016-la]. This is a different approach from the forensic statistical methods previously applied [@commissie2013], where results within one study are evaluated as opposed to other similar studies. This explains the discrepancy between the feasibility of evaluating the DECREASE trials by the statistical expert of the committee report and ours.

<!-- Hier evt nog wat vertellen over FOI? -->
<!-- Finally, we share our unsuccesful experiences in trying to further investigate the issue by requesting the DECREASE-VI raw data and call for more transparency. -->
To statistically investigate the evidence of data fabrication in the results of the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. First, we replicate the findings from a 2014 meta-analysis [@bouri2014] that contains sufficient information to estimate the deviation of the DECREASE-I and DECREASE-IV studies from the remaining studies. Second, we evaluate the veracity of the DECREASE-I and DECREASE-IV studies (i.e., the probability that these study findings occurred assuming no data fabrication occurred). Third, we reverse this assumption and assume that data fabrication did occur and estimate how many data points would have to be fabricated to reproduce the results of the DECREASE-I and DECREASE-IV studies if the remaining studies are regarded as estimating the true effect of beta-blockers on perioperative mortality.

# Step 1: reproducing meta-analysis of Bouri et al. (2014)

@@

To ensure that we used similar analysis procedures as in the 2014 meta-analysis [@bouri2014], we initially reproduced Bouri et al.'s estimates. This ensured that (1) their results are reproducible and (2) we are using the correct estimates in subsequent steps of our analyses. Using figures 2 and 3 from the original paper [@bouri2014], we extracted the raw event data for the 2 (control vs experimental) by 2 (event or no event) design, which we used to recompute the natural logarithm of the risk ratio and its standard error. The extracted event data is available at [osf.io/aykeh](https://osf.io/aykeh) and our analysis plan was preregistered at [osf.io/vnmzc](https://osf.io/vnmzc).

```{r, echo = FALSE}
suppressPackageStartupMessages(if(!require(metafor)){install.packages('metafor', repos = "http://cran.us.r-project.org")})
suppressPackageStartupMessages(library(metafor))
suppressPackageStartupMessages(if(!require(latex2exp)){install.packages('latex2exp', repos = "http://cran.us.r-project.org")})
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(library(knitr))

x12 <- '../'
# x12 <- ''
bouri_data <- read.csv(sprintf('%sdata/bouri_data.csv', x12))

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled these using the `R` package `metafor` [@viechtbauer2010]. We estimated a weighted random-effects model using the restricted maximum-likelihood estimator (i.e., `REML`) [@viechtbauer2005] to estimate the variance of effects. We used the default weighting procedure in the `metafor` package. When there was a zero-count for a cell, .5 was added to each cell of that trial, as is common in meta-analyses on risk- and odds ratios in order to deal with computational artefacts [@agresti2002]. The 2014 meta-analysis [@bouri2014] did not specify the effect variance estimate used; hence, minor discrepancies between our estimates and the original estimates could be due to differences in the estimation procedure.

@@

We analyzed the RRs to verify the original estimates (Figure 2 of the 2014 meta-analysis [@bouri2014]) and found that we were able to reproduce the estimates for the different sets of studies. Bouri et al. differentiated between the estimates from the non-DECREASE trials and the DECREASE trials. We confirmed the effect size estimates and the variance estimates for both the non-DECREASE- and the DECREASE trials, save for some minor discrepancies due to the estimation method. Table 1 depicts the original and reproduced values for both sets of studies.

```{r echo=FALSE,results='asis'}
table1 <- matrix('', ncol = 5, nrow = 5)
table1[1,3] <- 'Risk ratio'
table1[1,4] <- '$\\tau^2$'
table1[1,5] <- 'Confidence interval'
table1[2,1] <- 'Non-DECREASE'
table1[4,1] <- 'DECREASE'
table1[c(2,4),2] <- 'Original'
table1[c(3,5),2] <- 'Reproduced'
table1[2,3] <- 1.27
table1[2,4] <- 0.00
table1[2,5] <- '[1.01; 1.60]'

table1[3,3] <- round(exp(mod_clean$b), 2)
table1[3,4] <- round(mod_clean$tau2, 2)
table1[3,5] <- sprintf('[%s; %s]', round(exp(mod_clean$ci.lb), 2), round(exp(mod_clean$ci.ub), 2))

table1[4,3] <- 0.42
table1[4,4] <- 0.29
table1[4,5] <- '[0.15; 1.23]'

table1[5,3] <- round(exp(mod_dirty$b), 2)
table1[5,4] <- round(mod_dirty$tau2, 2)
table1[5,5] <- sprintf('[%s; %s]', round(exp(mod_dirty$ci.lb), 2), round(exp(mod_dirty$ci.ub), 2))

kable(table1, caption = 'The original- and reproduced meta-analytic results based on the data provided in the 2014 meta-analysis by Bouri et al.')
```

@@

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to replicate results presented in Figure 4 of the 2014 meta-analysis [@bouri2014]. Surprisingly, our results showed more evidence for unequal subgroups than the original meta-analysis [@bouri2014]; whereas the original test for subgroup differences was $\chi^2(1)=3.91,p=.05$, our analyses indicated that $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$. Additionally, the original analyses showed substantial heterogeneity ($I^2=74.4$\%) where we found an $I^2=`r round(mod_mods$I2, 3)`$\%. Different heterogeneity estimators (e.g., `DerSimonian-Laird` instead of `REML`) did not resolve this difference. We tried to clarify these discrepancies by e-mailing the original authors, but failed to receive a response.

<!--- Emailed bouri about this on 2016 03 16 -->

@@

Based on the non-DECREASE estimates from Step 1, we estimated the probability of obtaining the results in the DECREASE studies. To this end, we assumed that the non-DECREASE studies provide a valid representation of the true effect of beta-blockers on perioperative mortality (similar to Bouri et al. [@bouri2014]; but see the discussion). The estimated probability is also known as the veracity of the data [@peters2015], which indicates the probability of the observed data under a given true effect. We assumed that the non-DECREASE studies estimate the true population distribution of effects, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset (9 of 11 nonsignificant).

@@

Based on the estimated mean log RR and its credible interval in the non-DECREASE studies, we computed the probability of log RR in the DECREASE trials. The estimates of the non-DECREASE studies were obtained from Step 1, which include the estimated log RR (i.e., `r round(mod_clean$b, 2)`), and its 95\% credibility interval as provided by the package `metafor` (i.e., `r sprintf('[%s; %s]', round(mod_clean_pred$cr.lb, 3), round(mod_clean_pred$cr.ub, 3))`. We assumed a normal distribution of population effects with the estimated effect as the mean of the distribution. The 95\% credibility interval denotes the bounds of the normal distribution that covers 95\% of the density, where the standard deviation is calculated as the distance from the mean to either bound, divided by 1.96. This allows for an approximation of the population effect distribution, as depicted in Figure `r fig <- 1; fig`.

@@

pdf(sprintf('%sfigures/fig1.pdf', x12), width = 8, height = 6)
par(mar = c(.5,0,0,0))
plot(x = temp,
     dnorm(x = temp,
           mod_clean$b,
           (mod_clean_pred$pred - mod_clean_pred$cr.lb)/1.96),
     type = 'l',
     bty = 'n',
     xaxt = 'n',
     yaxt = 'n',
     xlab = '',
     ylab = '')

@@

invisible(dev.off())
```

```{r figure 1, fig.cap = 'Density plot of the estimated true effect distribution based on the non-DECREASE studies only, with the position of the DECREASE studies highlighted.', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig1.pdf", x12))
```

@@

Figure `r fig` indicates that the DECREASE trials are highly unlikely under the estimated effect distribution based on the non-DECREASE trials. More specifically, DECREASE-I (or more extreme) has a probability of $`r prob[1]`$ (less than 1 in a quintillion) and DECREASE-IV $`r prob[2]`$ (3 in a billion), which indicates they are unlikely to come from the same population effect distribution as the non-DECREASE trials. Observing two of such extremely unlikely results in this population effect distribution jointly is highly improbable, $`r prob[1]*prob[2]`$. This would therefore indicate that the DECREASE trials are severely different from the non-DECREASE studies.

Results from Step 1 indicated that no variance (i.e., homogeneity; $\tau^2=0$) of the effects was observed; given the small number of studies included (i.e., `r mod_clean$k`) this estimate is uncertain, however. We conducted sensitivity analyses in order to see how dependent results are on the specific heterogeneity estimate. The probability of observing both DECREASE trials stays approximately below 1 out of 1000 until the variance estimate is .25, and the probability stays approximately below 1 out of 100 until the variance estimate is .43 (see Figure `r fig <- fig + 1; fig`). Further research would be worthwhile to better estimate the variance in the effects of beta-blockers on perioperative mortality and what moderators affect the effect size (e.g., type of beta-blocker).

```{r, echo = FALSE}
pdf(sprintf('%sfigures/fig2.pdf', x12), width = 8, height = 6)
tau2 <- seq(0, 2, .01)

@@

par(mar = c(4, 4, 0, 0))
plot(tau2, prob_sens_t, type = 'l', xlab = TeX('$\\tau^2$'),
     ylab = 'p-value', bty = 'n')
points(x = tau2[c(26, 44)], y = prob_sens_t[c(26, 44)])
text(x = tau2[c(26, 44)],
     y = prob_sens_t[c(26, 44)] + .02,
     labels = c(TeX('$\\tau^2=.25$'), TeX('$\\tau^2=.43$')),
     cex = .8)
invisible(dev.off())
```
```{r figure 2, fig.cap = 'Sensitivity analyses for the p-value that indicates the probability of observing the results from the DECREASE studies, or more extreme results, based on the estimated true effect (non-DECREASE trials) and the accompanying variance estimate.', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig2.pdf", x12))
```

@@

We estimated the number of data points that would need to be fabricated to arrive at the estimates from the DECREASE studies, given that the non-DECREASE studies represent the true effect. This assumes, in contrast to Step 2, that the DECREASE studies in fact contain fabricated data. The estimates from Step 3 provide an indication of the extent of the potential data fabrication in the DECREASE studies under this assumption [@commissie2011;@commissie2012;@commissie2013;@bouri2014].

@@

In order to estimate the number of fabricated data points, we estimated the population effect distribution of mortality (in log odds) per condition in both the DECREASE- and non-DECREASE studies separately. For each of these, we ran a meta-analysis applying the same, verified methods used in Step 1. These separate analyses were necessary, considering that Step 1 only provided estimates of the overall risk ratio but not the odds conditional on exposure (beta-blocker or control) and study set (DECREASE- versus non-DECREASE). This resulted in four meta-analytic mortality estimates with an accompanying variance. Throughout the simulations, we use the point estimates to simulate genuine- and fabricated data, but supplement this by using the distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated data points in the DECREASE-I and DECREASE-IV studies [@casella2002]. With the inversion method, we iteratively hypothesized that *X* out of *N* data points were fabricated (i.e., $0, 1, ..., N$) in the DECREASE-I and DECREASE-IV studies separately. For each study, we simulated 10,000 datasets per hypothesis of *X* fabricated data points and *N-X* genuine data points. For each simulated dataset (exact simulation procedure in next paragraph), we determined the likelihood of the results with 
\begin{equation}
\label{eq1}
L(\theta|\pi_{E},\pi_{C})=\pi_{E}^{n_{11}}(1-\pi_{E})^{n_{12}} \times \pi_{C}^{n_{21}}(1 - \pi_{C})^{n_{22}}
\end{equation}
where $\pi_{E}$ indicates the mortality rate of the experimental condition as drawn from the meta-analytic effect distribution ($\pi_{C}$ indicates the same for the control condition). The likelihood was computed under both the fabricated effect estimates (i.e., $L_{fabricated}$) and the genuine data (i.e., $L_{genuine}$). Table 2 indicates which cell sizes the various $n_{XX}$ refer to within the (simulated) data. We compared the likelihoods to determine whether the simulated data were more likely to arise from the genuine- ($L_{genuine}$) or from the fabricated data ($L_{fabricated}$). Note that using the likelihoods for this is a minor deviation from the preregistration, where we initially planned on using $p$-value comparisons ([osf.io/vnmzc](https://osf.io/vnmzc)).

```{r echo=FALSE,results='asis'}
table2 <- matrix('', ncol = 3, nrow = 3)
table2[1,1] <- ''
table2[1,2] <- 'Dead'
table2[1,3] <- 'Alive'

table2[2,1] <- 'Beta-blockers'
table2[2,2] <- '$n_{11}$'
table2[2,3] <- '$n_{12}$'

table2[3,1] <- 'Control'
table2[3,2] <- '$n_{21}$'
table2[3,3] <- '$n_{22}$'

kable(table2, caption = 'Outcome possibilities within a simulated 2 (beta-blocker v control) by 2 (dead v alive) clinical trial.')
```

For each hypothesis of *X* out of *N* fabricated datapoints, we computed the probability that the fabricated data are more likely than the genuine data ($p_F=P(L_{fabricated}>L_{genuine})$).  Based on $p_F$, we computed the confidence interval for $X$ (i.e., $X_{LB};X_{UB}$). For a 95\% confidence interval, the lowerbound is equal to the $p_F$ closest to .025, whereas the upperbound is equal to the $p_F$ closest to .975.

We computed $p_F$ for all possibilities of *X* out of *N* fabricated datapoints in 10,000 randomly generated datsets, which were generated in three steps. For each dataset we:

1. Sampled (without replacement) *X* fictitious participants that would be the result of data fabrication (sampled across all conditions).

2. Determined the population mortality rate for each condition (i.e., for each cell as in Table 3). The meta-analytic point estimate was used or a population effect was randomly drawn from the meta-analytic effect distribution.

3. Simulated the number of deaths for the different conditions using a binomial distribution based on the mortality rate as determined in 2, resulting in the cell counts as in Table 3.

Based on the meta-analytic effect from 2 and the cell sizes from 3, we computed the  likelihoods of $L_{fabricated}$ and $L_{genuine}$ using Equation \ref{eq1}. As mentioned before, $p_F$ was computed as the probability that the data were more likely under the estimates resulting from the (allegedly) fabricated data (i.e., the DECREASE trials) than under the estimates resulting from the genuine data (i.e., the non-DECREASE trials; $p_F=P(L_{fabricated}>L_{genuine})$).

@@

# pdf(sprintf('%sfigures/fig3.pdf', x12), width = 8, height = 6)
# 
# par(mar = c(2.5, 5, 0.5, 0.5),
#     mfrow = c(2, 2))
# plot(res_i_uncertain,
#      xlab = '',
#      ylab = 'Proportion',
#      ylim = 0:1,
#      type = 'l',
#      lty = 1)
# lines(res_i_certain, pch = 2, type = 'l', lty = 2)
# abline(h = c(.025, .975), lty = 3)
# 
# legend(x = 0, y = .9, bty = 'n', lty = c(1, 2), legend = c("Distribution estimate", "Point estimate"))
# 
# plot(x = NA, y = NA,
#      xlim = c(0,1), ylim = c(0, length(res_i_certain)),
#      xlab = "Confidence in %", ylab = "Data points fabricated")
# 
# lines(x = seq(0, 1, .01), y = res_i_uncertain_lb)
# lines(x = seq(0, 1, .01), y = res_i_uncertain_ub)
# lines(x = seq(0, 1, .01), y = res_i_certain_lb, lty = 2, pch = 2, type = 'l')
# lines(x = seq(0, 1, .01), y = res_i_certain_ub, lty = 2, pch = 2, type = 'l')
# abline(v = .95, lty = 3)
# 
# # clip(x1 = 0, x2 = length(res_i_uncertain), y1 = 0.025, y2 = .975)
# # abline(v = which(abs(res_i_uncertain - .975) == min(abs(res_i_uncertain - .975))), lty = 1)
# # abline(v = which(abs(res_i_certain - .975) == min(abs(res_i_certain - .975))), lty = 2)
# 
# par(mar = c(4, 5, 0.5, 0.5))
# plot(res_iv_uncertain,
#      xlab = 'Data-points fabricated',
#      ylab = 'Proportion',
#      ylim = 0:1,
#      type = 'l',
#      lty = 1)
# lines(res_iv_certain, pch = 2, type = 'l',
#       lty = 2)
# abline(h = c(.025, .975), lty = 3)
# 
# plot(x = NA, y = NA,
#      xlim = c(0,1), ylim = c(0, length(res_iv_certain)),
#      xlab = "Confidence in %", ylab = "Data points fabricated")
# 
# lines(x = seq(0, 1, .01), y = res_iv_uncertain_lb)
# lines(x = seq(0, 1, .01), y = res_iv_uncertain_ub)
# lines(x = seq(0, 1, .01), y = res_iv_certain_lb, lty = 2, pch = 2, type = 'l')
# lines(x = seq(0, 1, .01), y = res_iv_certain_ub, lty = 2, pch = 2, type = 'l')
# abline(v = .95, lty = 3)
# invisible(dev.off())

pdf(sprintf('%sfigures/fig3.pdf', x12), width = 8, height = 6)

par(mar = c(5, 5, 2.5, 0),
    mfrow = c(2, 2))
plot(res_i_uncertain,
     xlab = 'Data points fabricated',
     ylab = 'Proportion',
     main = 'DECREASE I',
     ylim = 0:1,
     type = 'l',
     lty = 1)
lines(res_i_certain, pch = 2, type = 'l', lty = 2)
abline(h = c(.025, .975), lty = 3)

@@

par(mar = c(5, 0, 2.5, 0.5))

plot(res_iv_uncertain,
     xlab = 'Data-points fabricated',
     ylab = '',
     main = 'DECREASE IV',
     ylim = 0:1,
     yaxt = 'n',
     type = 'l',
     lty = 1)
lines(res_iv_certain, pch = 2, type = 'l',
      lty = 2)
abline(h = c(.025, .975), lty = 3)

par(mar = c(4, 5, 0.5, 0))

@@

par(mar = c(4, 0, 0.5, 0.5))

plot(x = NA, y = NA,
     xlim = c(0,1), ylim = c(0, length(res_iv_certain)),
     xlab = "Confidence in %", ylab = "",
     yaxt = 'n')

lines(x = seq(0, 1, .01), y = res_iv_uncertain_lb)
lines(x = seq(0, 1, .01), y = res_iv_uncertain_ub)
lines(x = seq(0, 1, .01), y = res_iv_certain_lb, lty = 2, pch = 2, type = 'l')
lines(x = seq(0, 1, .01), y = res_iv_certain_ub, lty = 2, pch = 2, type = 'l')
abline(v = .95, lty = 3)
invisible(dev.off())
```

```{r figure 3, fig.cap = 'Test', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig3.pdf", x12))
```

```{r}
res_i_certain
res_i_uncertain
res_iv_certain
res_iv_uncertain
```
